---
title: Computing numerical summaries and parametric distributions
subtitle: BDSI R Training I
format:
  anu-light-revealjs:
    width: 1920
    height: 1080
    auto-stretch: false
    html-math-method: katex
    self-contained: true
    css: assets/custom.css
    footer: anu-bdsi.github.io/training-R1/
author: Emi Tanaka
institute: Biological Data Science Institute
date: 2023/11/02
date-format: "D[nd] MMMM YYYY"
execute:
  echo: true
---

## Functions in R 

::: incremental

- There are many functions in R!
- Generally if you need to compute some numerical summary that is common in your field, then there is probably already an existing function in the R ecosystem.
- Try always searching it on a search engine (e.g. Google) with the right keywords. 
- If it's computed from a community contributed R package, then check to see if there's some quality indicators:
  - Is it actively maintained?
  - Is it widely used?
  - Does the package have tests for its functions? Etc.

:::

## Base packages 

::: incremental

- R has 7 packages (`stats`, `graphics`, `grDevices`, `utils`, `datasets`, `methods`, `base`), collectively referred to as the "base packages", that are loaded automatically when you launch it. 
- The functions in the base packages are generally well-tested and trustworthy.

:::


## Artihmetics

- Many of the arithmetic functions come from `base`.
- You can see `library(help = "base")` for indexed help files.

```{r}
#| echo: true
sqrt(3)
abs(-3)
exp(1)
log(4, base = exp(1))
sum(1:3)
```


## Numerical summaries 

- Numerical summaries generally come `base` or `stats` package.
- Some common numerical summaries include:
  - Mean: `mean()`
  - Median: `median()`
  - Five number summary: `fivenum()`
  - Minimum: `min()`
  - Maximum: `max()`
  - Quantile: `quantile()`
  - Correlation coefficient: `cor()`


## Missing values 

::: incremental

- `NA` in R denotes missing values -- there are in fact different types of missing values (`NA_character_`, `NA_integer_`, `NA_real_`, `NA_complex_`).

:::

. . . 

- When there are missing values, it can cause issues in the computation.

```{r}
x <- c(2.3, NA, 4.7)
mean(x)
```

. . . 

- Below we remove the missing values:

```{r}
mean(x, na.rm = TRUE)
```

. . . 

- Notice that the above is different to below when there are missing value(s):

```{r}
sum(x, na.rm = TRUE) / length(x)
```


## Some parametric distributions 

- The density (`d`), distribution (`p`) or quantile (`q`) functions of a parametric distribution are generally in the `stats` package.
- There are functions to generate random values from a particular parametric distribution (`r`).

- Some examples are:

::: flex

::: {.w-25}

Normal distribution

- `dnorm()`
- `pnorm()`
- `qnorm()`
- `rnorm()`


:::

::: {.w-25 .pl3}

t-distribution 

- `dt()`
- `pt()`
- `qt()`
- `rt()`

:::

::: {.w-25 .pl3}

Poisson distribution 

- `dpois()`
- `ppois()`
- `qpois()`
- `rpois()`

:::

::: {.w-25 .pl3}

F distribution 

- `df()`
- `pf()`
- `qf()`
- `rf()`

:::

:::




