---
title: Basics of R Programming 
subtitle: BDSI R Training I
format:
  anu-light-revealjs:
    width: 1920
    height: 1080
    auto-stretch: false
    html-math-method: katex
    self-contained: true
    css: assets/custom.css
    footer: anu-bdsi.github.io/training-R1/
author: Emi Tanaka
institute: Biological Data Science Institute
date: 2023/11/02
date-format: "D[nd] MMMM YYYY"
execute: 
  echo: true
---

## What is R?

::: incremental

- **R** is a programming language predominately for data analysis
- **RStudio Desktop** is an integrated development environment (**IDE**) that helps you to use R
- **R packages** are community developed extensions to R [(much like apps on your mobile)]{.grey}
- The Comprehensive R Archive Network (**CRAN**) is a volunteer maintained repository that hosts submitted R packages that are approved [(much like an app store)]{.grey}
  - There are over 10,000 packages available
  - The qualities of R packages vary 
- There are other repositories that host R packages, e.g. **Bioconductor** for bioinformatics [(we won't cover this)]{.grey}
:::

## Simple algebraic operations

- You can use R like a calculator

```{r}
#| output-location: fragment
3 + 2 * (2 - 6 / 3) 
```

. . . 

- Assignments to object:
  - You can assign values to objects using `<-` or `=`
  - The name of the object can be variable so long as it is _syntactically valid_ (no spaces and most special characters, and the name cannot start with a digit)

```{r}
a <- 3
b = 2
2 * a - b 
```

## Vectors


