---
title: Data wrangling with dplyr
subtitle: BDSI R Training I
format:
  anu-light-revealjs:
    width: 1920
    height: 1080
    auto-stretch: false
    html-math-method: katex
    self-contained: true
    css: assets/custom.css
    footer: anu-bdsi.github.io/training-R1/
author: Emi Tanaka
institute: Biological Data Science Institute
date: 2023/11/03
date-format: "D[rd] MMMM YYYY"
execute: 
  echo: true
---



## A grammar of data manipulation

```{r}
library(dplyr) # or library(tidyverse)
```

```{r}
#| echo: false
options(width = 60)
```


* `dplyr` is a core package in `tidyverse` 
* The earlier concept of `dplyr` (first on CRAN in 2014-01-29) was implemented in `plyr` (first on CRAN in 2008-10-08)
* The functions in `dplyr` has been evolving frequently but `dplyr` v1.0.0 was released on CRAN in 2020-05-29
* This new version contained new "verbs"
* The major release suggests that functions in `dplyr` are maturing and thus the user interface is unlikely to change


::: {.absolute bottom=0 .f3}

Hadley Wickham, Romain François, Lionel Henry and Kirill Müller (2020). dplyr: A Grammar of Data Manipulation. R package version 1.0.2. 
  
Hadley Wickham (2011). The Split-Apply-Combine Strategy for Data Analysis. Journal of Statistical Software, 40(1), 1-29.
  
:::




## `dplyr` "verbs"


* The main functions of `dplyr` include: 
<center>
<table style="width:60%">
  <tr>
    <td><code>arrange</code></td>
    <td><code>select</code></td>
    <td><code>mutate</code></td>
    <td><code>rename</code></td>
    <td><code>group_by</code></td>
    <td><code>summarise</code></td>
  </tr>
</table>
</center>

* Notice that these functions are _verbs_

. . . 

* Functions in `dplyr` generally have the form:

<center>

`verb(data, args)`

</center>

* I.e., the first argument `data` is a `data.frame` object

. . . 

* What do you think the following will do?

<center>

`rename(mtcars, miles_per_gallon = mpg)`  
`arrange(mtcars, wt)`

</center>

## Pipe operator %>%

* Almost all tidyverse packages import the `magrittr` package to use `%>%`

* `x %>% f(y)` is the same as `f(x, y)`
* `x %>% f(y) %>% g(z)` is the same as `g(f(x, y), z)`
* When you see the pipe operator `%>%`, read it as "and then"

```{r}
mtcars %>%                          # take mtcars data, and then
 rename(miles_per_gallon = mpg) %>% # rename mpg as miles_per_gallon, and then
 arrange(wt)                        # arrange row by wt
```



::: {.absolute bottom=0 .f3}

Stefan Milton Bache and Hadley Wickham (2020). magrittr: A
  Forward-Pipe Operator for R. R package version 2.0.1.

:::


## Subsetting by column

::: flex

::: {.w-50}

```{r, eval = FALSE}
select(mtcars, c(mpg, cyl))
select(mtcars, c("mpg", "cyl"))
select(mtcars, mpg, cyl)
select(mtcars, "mpg", "cyl")
```
All the same result as below
```{r}
mtcars %>% 
  select(mpg, cyl)
```

:::

::: {.w-50 .pl3}

```{r}
mtcars %>% select(mpg)
```

- Selecting one column doesn't "drop" it to a vector.
- If you really want the vector then use `pull(mpg)`.

:::

:::


## Subsetting by row 

```{r}
mtcars %>% 
  slice(3:1)
```

```{r}
mtcars %>% 
  filter(rownames(.) %in% c("Datsun 710", "Mazda RX4"))
```
* Use `slice` to subset by *index* and `filter` by *logical vector*


<div class="ba" style="position:absolute;right:20px;margin-right:0px!important;top:50px;margin-left:0;width:900px;font-size: 20pt;background-color:white;">
<ul>
<li>Placeholder binding</li>
<ul>
<li><code>x %>% f(y, g(.))</code> is the same as <code>f(x, y, g(x))</code></li>
<li><code>x %>% f(y, .)</code> is the same as <code>f(y, x)</code></li>

<li><i class="fas fa-exclamation-triangle red"></i> Note: row names do not follow tidy data principles</li>

<li>Use <code class="monash-blue">tibble::rownames_to_column()</code> to convert rownames to a column to make into a tidy data</li>
</ul>
</div>



## Adding or modifying a column 

```{r mutate}
mtcars %>% 
  mutate(gpm = 1 / mpg,
          wt = gpm^2,
          wt = if_else(cyl==6, 10, wt),
          hp = case_when(cyl==6 ~ 11,
                         cyl==4 ~ 10,
                           TRUE ~ 3))
```


<div class="ba" style="position:absolute;right:20px;margin-right:0px!important;top:50px;margin-left:0;width:550px;font-size: 20pt;background-color:white;">
<ul>
<li>Evaluation in <code>mutate</code> is done sequentially based on input order</li>
<li>So you refer to the newly created variable in later input</li>

<li>You can call multiple <code>mutate</code> but computational performance is usually better if done within the same <code>mutate</code> call</li>
<pre>
<code>mtcars %>%
  mutate(gpm = 1 / mpg) %>% 
  mutate(wt = gpm^2)
</code></pre>
</ul>
</div>





## Sorting columns 


```{r}
mtcars %>% 
  select(sort(names(.)))
```

```{r}
mtcars %>% 
  relocate(wt, gear, .after = mpg)
```



## Calculating statistical summaries by group 


::: flex

::: {.w-50}

🎯 Calculate the _average_ weight (`wt`) of a car for each gear type in (`gear`)  `mtcars`

```{r}
mtcars %>% 
  group_by(gear) %>% 
  summarise(avg_wt = mean(wt))
```

:::

::: {.w-50 .pl3}

🎯 Calculate the _median_ weight (`wt`) of a car for each gear (`gear`) and engine (`vs`) type in  `mtcars`

```{r}
mtcars %>% 
  group_by(gear, vs) %>% 
  summarise(avg_wt = mean(wt),
            med_wt = median(wt))
```

:::

:::



##  `across` 

* Using `across`, you can more easily apply a function to multiple columns

```{r, message = FALSE}
#| code-line-numbers: 3
mtcars %>% 
  group_by(gear, vs) %>% 
  summarise(across(everything(), mean)) 
```

##  `where` 

* You can combine `across` with the selection helper `where` 

```{r, message = FALSE}
#| code-line-numbers: 3
mtcars %>% 
  group_by(gear, vs) %>% 
  summarise(across(where(function(x) n_distinct(x) > 10), mean)) 
```


##  `c_across`

* Remember tidy selection only works with functions that are compatible

```{r, message = FALSE}
#| code-line-numbers: 4
mtcars %>% 
  rowwise() %>% 
  summarise(disp = disp, hp = hp, drat = drat, wt = wt,
            score = sum(c_across(disp:wt))) 
```

##  `rowwise`?

* What happens if you omit `rowwise`?

```{r, message = FALSE}
#| code-line-numbers: 4
mtcars %>% 
  #rowwise() %>% 
  summarise(disp = disp, hp = hp, drat = drat, wt = wt,
            score = sum(c_across(disp:wt))) 
```




## Lazy and non-standard evaluation

* In Base R:

```{r}
#| echo: false
options(width = 80)
```


```{r}
subset(mtcars, mpg > 31)
```

. . . 

* But the second argument cannot be evaluated:

```{r, error = TRUE}
mpg > 31
```

. . . 

* R employs what is called **lazy evaluation** for function inputs 

* **Non-standard evaluation** uses this feature to capture the input expression within the function and evaluate only when requested



## Tidy evaluation [Part 1]{.f2}

* Tidy evaluation builds on the lazy and non-standard evaluation and is implemented in `rlang`
* All core tidyverse packages import `rlang`
* So what does it do?

. . . 

* Let's consider `filter`, the Tidyverse version of `subset`

```{r}
filter(mtcars, mpg > 31)
```

* <i class="fas fa-exclamation-triangle animated flash red"></i> If you get an error using `filter`, replace it with `dplyr::filter`<br> [for those interested, `dplyr::filter` is a conflict with `stats::filter` and it may be using `stats::filter` instead... I've fallen into this trap so many times!]{.f2}


::: {.absolute bottom=0 .f3}

Lionel Henry and Hadley Wickham (2020). rlang: Functions for Base
  Types and Core R and 'Tidyverse' Features. R package version
  0.4.8.

:::


## Tidy evaluation [Part 2]{.f2}

* Suppose we have a silly function that subsets `mtcars` for a given condition

```{r}
myCarSubset <- function(cond) subset(mtcars, cond)
myCarFilter <- function(cond) filter(mtcars, cond)
```

. . . 

* This causes an issue because `cond` is evaluated before it is parsed into `subset` or `filter`

```{r, error = TRUE}
myCarSubset(mpg > 31)
myCarFilter(mpg > 31)
```


## Tidy evaluation [Part 3]{.f2}

* Functions that use non-standard evaluation is problematic

```{r curly}
myCarSubsetNew <- function(cond) subset(mtcars, {{ cond }})
myCarFilterNew <- function(cond) filter(mtcars, {{ cond }})
```
. . . 

```{r, error = TRUE, eval = TRUE}
myCarSubsetNew(mpg > 31)
```

. . . 

```{r, error = TRUE, eval = TRUE}
myCarFilterNew(mpg > 31)
```


* `{{{ }}}` only works if the underlying function implements `rlang`

## Data masking [Part 1]{.f2}


```{r, error = TRUE}
ind <- 1:nrow(cars) # nrow(cars) = 50
filter(cars, ind > 49)
```


* For any variables that don't exist in the data, R searches the parental environment for evaluation.

. . . 

```{r}
speed <- c(40, 51)
filter(cars, speed > 24)
```


* The variables in data take priority for reference over those in parental environment



## Data masking [Part 2]{.f2}

::: flex

::: {.w-50}

```{r data-mask, eval = TRUE, echo = TRUE}
speed <- 1:nrow(cars)
filter(cars, .data$speed > 24)
filter(cars, .env$speed > 24)
```


:::

::: {.w-50 .pl3}

* In Tidyverse, you can be explicit whether the variable is in the data or in the parental environment
* `.data` is a special pronoun referring to variables in the data parsed in the first argument
* `.env` is a special pronoun referring to variables in the environment (i.e. _not_ in the data parsed in the first argument)

:::

:::


## Tidy select [Part 1]{.f2}

* Tidyverse packages generally use syntax from the `tidyselect` package for variable/column selection

```{r}
immunity <- readr::read_csv("https://github.com/anu-bdsi/training-R1/raw/main/data/marcato-2021-immunity.csv")
str(immunity) 
```

## Tidy select [Part 2]{.f2}

::: flex

::: {.w-50}

```{r select}
immunity %>% 
 select(Timepoint, Batch, Animal)
```


:::

::: {.w-50 .pl3}


```{r select2}
immunity %>% 
  select(Timepoint:Animal)
```


:::

:::


- The `tidyselect` syntax `:` can be used to select contiguous columns in the data.


## Tidy select [Part 3]{.f2}


```{r}
immunity %>% 
  select(contains("perf"))
```

```{r}
immunity %>% 
  select(starts_with("Titer"))
```



## Selection language [Part 1]{.f2}

* `:` for selecting contiguous variables
* `!` for taking complement set of variables
* `&` or `|` for selecting intersection or union of two sets of variables, e.g. 
```{r}
immunity %>% 
  select(starts_with("Titer") & ends_with("IgG")) %>% 
  str()
```

* `c()` for combining selections
* `everything()` to select all variables
* `last_col()` to select last variable, with option of an offset


## Selection language [Part 2]{.f2}


* `starts_with()` selects columns with the given prefix
* `ends_with()` selects columns with the given suffix
* `contains()` selects columns with a literal string
* `matches()` selects columns that match the regular expression
* `num_range()` selects columns with a numerical range
* `all_of()` for selecting columns based on a character vector
* `any_of()` is the same as `all_of()` but no error when variables do not exist
* `where()` selects based on where given function return TRUE

```{r, eval = FALSE}
help(language, package = "tidyselect")
```


